---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "07/04/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)
sce <- DropletUtils::read10xCounts(
  samples = c("raw" = "/project/obds/shared/resources/4_r_single_cell/pbmc_5k/raw_feature_bc_matrix/"), col.names = TRUE)
  
  
sce
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
library(tidyverse)
colSums(counts(sce)) %>% head()
colSums(counts(sce)) > 0 %>% head()
#which(colSums(counts(sce)) >0 ) also works
sce <- sce[, colSums(assay(sce, "counts")) > 0]
sce
```
```{r}
rank_df<-DropletUtils::barcodeRanks(sce)
head(rank_df)



ggplot(as.data.frame(rank_df),aes(x=rank, y=total))+geom_point()+scale_x_log10()+scale_y_log10()+geom_hline(yintercept = 100,colour="red")
```


- Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(123)
out <- DropletUtils::emptyDrops(assay(sce, "counts"))
out
#everything is less than 100UMI, the columns will be NA have not been tested
summary(out$Total >100)
subset(out, Total >100) %>% head()
ggplot(as.data.frame(out), aes(x=FDR))+geom_histogram()
```

**How do you read the output? Which droplets would you consider empty?**

> Answer:
> 
> The output is a table with a row for each GEM barcode and columns for various pieces of information.
> - `Total` represents the total UMI count.
> - `FDR` is the corrected p-value indicating the probability that a droplet is empty (i.e., originate from the null distribution).

>
> 
- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}
out$empty <- out$FDR<0.001
head(out)
which(out$empty == TRUE)
##return the location of the cells with FDR<0.001
sce_filtered_emptydrops <- sce[,which(out$empty ==TRUE)]
sce_filtered_emptydrops
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
plot_data <- tibble(
  library_size = colSums(assay(sce, "counts")),
  empty_droplet = out$FDR > 0.01
)
ggplot(plot_data, aes(empty_droplet, library_size)) +
  geom_jitter(width = 0.1, height = 0) +
  scale_y_continuous(trans = "log10") +
  cowplot::theme_cowplot()
  
```

```{r}
is.mito <- grepl("^MT-", rowData(sce)$Symbol)
library(scuttle)
sce <- scuttle::addPerCellQC(sce, percent_top = 50, subsets = list(MT=is.mito))

colData(sce)[["empty"]] <- out$empty
#add a empty column true for cells, false for empty droplets 
plot1 <- colData(sce) %>%
  as_tibble() %>%
  ggplot() + 
  geom_violin(aes(empty, sum), scale = "width") + 
  labs(x="Is cell containing droplet", y = "Total UMI")
plot2 <- colData(sce) %>%
  as_tibble() %>%
  ggplot() + 
  geom_violin(aes(empty, detected), scale = "width") + 
  labs(x="Is cell containing droplet", y = "features")
plot3 <- colData(sce) %>%
  as_tibble() %>%
  ggplot() + 
  geom_violin(aes(empty, subsets_MT_percent), scale = "width") + 
  labs(x="Is cell containing droplet", y = "MT percentage")
cowplot::plot_grid(plot1, plot2, plot3, nrow=1)
  
  
```




```{r}
sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- 
  
  
all_barcodes <- colnames(sce)
compare_table <- tibble(
  emptydrops = all_barcodes %in%    ,
  cellranger = all_barcodes %in%    ,
)
table(compare_table)
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

- Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(123)
sce_scdblfinder <- scDblFinder(
  sce = sce_filtered_emptydrops,
  clusters = TRUE,
)
table(colData(sce_scdblfinder)$scDblFinder.class)
plotDoubletMap(sce_scdblfinder)
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer:
> 
> 
> 
> 
- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.
  
  
```{r}
#filter for singlets
sce_filtered_scdblfinder <- sce_scdblfinder[, sce_scdblfinder$scDblFinder.class == "singlet"]
sce_filtered_scdblfinder
```

## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
# find mitochondrial genes
is.mito <- grep("^MT-", rowData(sce_scdblfinder)$Symbol)



# subset based on mitochondrial genes
library(scuttle)



sce_scdblfinder <- scuttle::addPerCellQC(
x = sce_scdblfinder,
subset = list(MT = is.mito))



# view new object
colData(sce_scdblfinder)



# create plots



# plot to show UMI values for singlets and doublets
plotA <- colData(sce_scdblfinder) %>%
as_tibble() %>%
ggplot() +
geom_violin(aes(x = scDblFinder.class, y = sum), scale = "width") +
labs(x = "cell class", y = "total UMI")



# plot to show features for singlets and doublets
plotB <- colData(sce_scdblfinder) %>%
as_tibble() %>%
ggplot() +
geom_violin(aes(x = scDblFinder.class, y = detected), scale = "width") +
labs(x = "cell class", y = "features")



# plot to show % mito for singlets and doublets
plotC <- colData(sce_scdblfinder) %>%
as_tibble() %>%
ggplot() +
geom_violin(aes(x = scDblFinder.class, y = subsets_MT_percent), scale = "width") +
labs(x = "cell class", y = "MT percent")



cowplot::plot_grid(plotA, plotB, plotC, nrow = 1)
  
```

```{r}
is.mt <- 
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}
sce_scdblfinder <- 
sce_scdblfinder <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 
sweep.res.list <- 
sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}
sce_filtered_scdblfinder
sce_decontX <- decontX(sce_filtered_scdblfinder, varGenes =2000)
print(sce_decontX)
#check new col created
colnames(colData(sce_decontX))
#visualise 
colData(sce_decontX) %>%
  as_tibble() %>%
  ggplot(aes(decontX_contamination)) +
  geom_histogram()

ggplot(as.data.frame(colData(sce_decontX)), aes(decontX_contamination))+ geom_histogram()
```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
umap <- reducedDim(sce_decontX, "decontX_UMAP")
plotDimReduceCluster(x = sce_decontX$decontX_clusters, dim1 = umap[,1], dim2 = umap[,2])
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> Cluster 5
> 
- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}

sce_decontX <- logNormCounts(sce_decontX, assay.type = "decontXcounts")
rownames(sce_decontX) <- uniquifyFeatureNames(ID=rowData(sce_decontX)$ID, names = rowData(sce_decontX)$Symbol)

plotDimReduceFeature(assay(sce_decontX, "logcounts"), features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1"),dim1 = umap [ ,1], dim2 = umap [ ,2])
  
```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}
?plotDecontXMarkerPercentage
markers <- list( Tcell_Markers = c("CD3E", "CD3D"), Bcell_Markers = c("CD79A", "CD79B", "MS4A1"), Monocyte_Markers = c("S100A8", "S100A9", "LYZ"), NKcell_Markers = "GNLY" )


cellTypeMappings <- list(
  "T-cells" = c(1,3), "B-cells" = 4, "Monocytes" = c(2,10), "NK cells" = c(6,7)
)
  
  
plotDecontXMarkerPercentage(x = sce_decontX, markers = markers, groupClusters = cellTypeMappings)
  
```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  
  
  
  
)
```

- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r}
plotDecontXMarkerExpression(x = sce_decontX, markers = markers, groupClusters = cellTypeMappings
)


```