```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(tidyverse)
library(patchwork)
library(DT)
library(gprofiler2)
```

# Exercise

## Read in the RDS object we created and save from Seurat day 1

- use the `readRDS()` function to read in previously saved object

```{r}
seurat_after_qc <- readRDS("seurat_after_qc.RDS")
```

- Check the Default Assay

```{r}
DefaultAssay(seurat_after_qc)
```


## Apply SCTransfrom  normalisation

Use `SCTransform()` function

SCTransform vignette: https://satijalab.org/seurat/articles/sctransform_vignette.html 

`SCTransform()`  command replaces `NormalizeData()`, `ScaleData` and `FindVariableFeatures()` run for the RNA assay in day 1 Seurat

Should we remove any confounding variables like we did for the RNA assay for Day 1?

Do we want to use the same number of variable featuresn(n=1000) or more than what we used for `NormalizeData()` function.


```{r}
?SCTransform

head(seurat_after_qc)
seurat_after_qc <- SCTransform(
  object = seurat_after_qc,
  assay = "RNA",
  new.assay.name = "SCT", 
  do.correct.umi = TRUE,
  ncells = 5000, 
  variable.features.n = 3000, 
  vars.to.regress = "percent_mt",
  return.only.var.genes = TRUE)

#don't use UMI in vars.to.regress as function does this already
```
where is the new normalisation stored?
Answer: 

explore the `seurat_after_qc` objects meta.data and assays ? 


```{r}
Assays(seurat_after_qc)

head(seurat_after_qc)
seurat_after_qc
dim(seurat_after_qc@assays$RNA)
dim(seurat_after_qc@assays$SCT)
```

```{r}
?SCTransform
```

is there a change? 
Answer: yes, new Assay called SCT 

Are there new columns in the meta.data?
Answer: yes, recalcultaed library size based on corrected depth. 


# Exercise

## Visulaisation

- The library size and number of features detected per cell is already present in the Seurat object.

- When you run `SCTransform` you get two new variables for library size and features for SCT normalisation

 - Use the function `VlnPlot()` to display compare RNA assay vs SCT assay library size in one plot 
 
 - Vlnplot of features detected in RNA vs SCT  in a single violin plot.

- check to see how you can have fixed ylims in the `VlnPlot()` function

```{r}
head(seurat_after_qc)
VlnPlot(seurat_after_qc,features = c("nCount_RNA", "nCount_SCT"), same.y.lims = TRUE )
VlnPlot(seurat_after_qc,features = c("nCount_RNA", "nCount_SCT"), same.y.lims = FALSE )

```




## Bonus-
- Visualise the distribution of SCT corrected count data for a feature of your choice.
- Visualise the distribution of SCT logs counts in the dataa slot for a feature (e.g. LYZ)

- Visualise the distribution of SCT scale.data count feature of your choice. 

lets choose LYZ like day 1

```{r}
DefaultAssay(seurat_after_qc) <- "SCT"
head(seurat_after_qc)

ggplot_lyz_corrected <- ggplot(FetchData(seurat_after_qc, vars = "LYZ", slot = "counts"), aes(LYZ)) +
    geom_histogram(fill = "grey", color = "black", bins = 100) +
    coord_cartesian(ylim = c(0, 500)) +
    cowplot::theme_cowplot()
ggplot_lyz_normalised <- ggplot(FetchData(seurat_after_qc, vars = "LYZ", slot = "data"), aes(LYZ)) +
    geom_histogram(fill = "grey", color = "black", bins = 100) +
    coord_cartesian(ylim = c(0, 500)) +
    cowplot::theme_cowplot()
ggplot_lyz_scaled <- ggplot(FetchData(seurat_after_qc,vars = "LYZ", slot = "scale.data" ), aes(LYZ)) +
    geom_histogram(fill = "grey", color = "black", bins = 100) +
    coord_cartesian(ylim = c(0, 500)) +
    cowplot::theme_cowplot()
cowplot::plot_grid(ggplot_lyz_corrected, ggplot_lyz_normalised,ggplot_lyz_scaled, ncol = 1)
```
Use the function `VariableFeaturs` to pull out the 1:10 the variable genes after SCT and compare to 1:10 from the RNA assay

do we need to change any arguments to get the variables genes specific to the SCT or RNA assay

```{r}
?VariableFeatures.Assay
VariableFeatures(seurat_after_qc, assay = "RNA")[1:10]
VariableFeatures(seurat_after_qc, assay = "SCT")[1:10]
```





# Exercise


## Dimensionality reduction on SCT transformed data

- Run a principal component analysis and UMAP on the Seurat object.
- Check the Default assay

- Do we want to change the `reduction.name` argument so that we can still keep the RNA assay based PCA results?
  
```{r}
DefaultAssay(seurat_after_qc) <- "SCT"
seurat_after_qc <- RunPCA(seurat_after_qc, reduction.name = "sct.pca" )
```

check to see what reductions are now present in the object

```{r}
Reductions(seurat_after_qc)
?RunPCA
```

- First, visualise the amount of variance explained the top principal components for SCTransformed data (number of your choice).
  How many principal components would you use for downstream analyses?
  
- do we need to specificy the correction reduction?

```{r}
?ElbowPlot

ElbowPlot( seurat_after_qc, ndims = 100 , reduction = "sct.pca" )
```

- Run the UMAP technique on the first 20 principal components stored in the Seurat object and visualise the result as a scatterplot.

how can we change the reduction name from defauult "umap" to "sct.umap"

how can we specify that we want to use pca run on the SCT Assay (sct.pca) in the previous step?



```{r}
?RunUMAP
seurat_after_qc <- RunUMAP(seurat_after_qc, dims = 1:20, reduction ="sct.pca", reduction.name = "sct.umap")
```

use DimPlot() to plot the umap, what happens if you try to specify different reductions with UMAPPlot()?

compare RNA based umap with sct.umap

```{r, fig.width=10}
p1 <- DimPlot(seurat_after_qc, reduction ="umap")   + ggtitle("RNA UMAP")
p2 <-   DimPlot(seurat_after_qc, reduction = "sct.umap") + ggtitle("SCT UMAP")
 p1 + p2
```


# Exercise

## Clustering on SCTransformed data



- Then, compute the graph of nearest neighbours using the function `FindNeighbors()`.
  Which principal components are used by default?
  Instead, specify the number of components that you have chosen.
  Have you chosen the right `reduction`  and `assay`? 
  
```{r}
seurat_after_qc <-  FindNeighbors(seurat_after_qc, dims = 1:20,assay = "SCT",reduction = "sct.pca")
```


- Finally, compute cluster labels.
  What is the default setting for the `resolution` argument?
  Instead, set it to `0.5`.
  Do you expect more or fewer clusters following that change?
  What other parameters would you also try to experiment with?
  - should we specify the `graph.name` ?

```{r}
seurat_after_qc <- FindClusters(seurat_after_qc, resolution = 0.5, graph.name = "SCT_snn")
```


Check cluster assigment between SCT and RNA workflow

if you use the same resolution = o.5 and dims as RNA workflow do you get the same number of cluster or more?

are cells in the same cluster across both RNA and SCT
```{r}
table(seurat_after_qc$RNA_snn_res.0.5, seurat_after_qc$SCT_snn_res.0.5)

```

- Visualise the SCT cluster labels on the SCT transformed UMAP scatter plot and the RNA cluster labels on the RNA umap
  How would you describe the agreement between the UMAP layout and the clustering results?= for SCT vs RNA Assay results
  

```{r}
p1 <- DimPlot(seurat_after_qc, reduction = "umap", group.by = "RNA_snn_res.0.5") + ggtitle("RNA UMAP")
p2 <- DimPlot(seurat_after_qc, reduction = "sct.umap", group.by = "SCT_snn_res.0.5" ) + ggtitle("SCT UMAP")
p1 + p2
```


Plot some known cell-type markers for PBMC datasets, does the SCT better seperate the celltypes?

CD14+ Monocyte : LYZ, CD14
CD16 Monocytes : FCGR3A, MS4A7

CD4 T :  CD4, IL76

CD8 T : CD8A, CD3D

NK : GNLY, GZMB,NKG7

B Cell :  MS4A1 , CD79A

DC : CST3, FCER1A

Platelets : PPBP


```{r, fig.width=10, fig.height=8}
FeaturePlot(seurat_after_qc, features = c("LYZ","FCGR3A","CD4","CD8A","GNLY","MS4A1","CST3","PPBP"), reduction = "sct.umap")
FeaturePlot(seurat_after_qc, features = c("LYZ","FCGR3A","CD4","CD8A","GNLY","MS4A1","CST3","PPBP"), reduction = "umap")
```



# Bonus exercise to try in your own time:: Pathway analysis on Cluster markers for all clusters


- Choose either RNA workflow based generated or SCT generated based  seurat marker results

- we will be using Gprofiler `gost()` function fot multiple gene lists at the same time

- First we need to  filter seurat get top significant (p_adj < 0.05)  upregulated genes with a LogFC threshold (decided by you) for each cluster , use `dplyr:: group_by()` and `dplyr::filter` to get gene list for each cluster
then only select cluster and gene

- We then use ` split()`  on the filtered_df to divide gene markers into list  of multiple character vectors containing genes split by cluster

- can refer to the pahtway analysis code from week 3 but for h.sapiens

```{r}

```

 We then run pathway analysis using `gost()` with multi_query = TRUE
 
 - to generate all_genes_id , we use all genes present in either the RNA assay or SCT assay , as we have already filtered out lowly expressed genes not present in certain cells .
 
```{r}
# Choose Default assay based on if running pathway analyses on RNA or SCT results
DefaultAssay(seurat_after_qc) <- ""
# create a vector of  of all genes 
all_genes_id <- rownames(seurat_after_qc)
multi_gostquery_results_obj <- gost(query = seurat_clusters_list,
                                    organism = "hsapiens", 
                                    exclude_iea = TRUE, # change to True as have alot of results otherwise
                                    correction_method = "g_SCS",  # this is thier own multiple testing coorrection - actually more strignent then bonferroni or BH 
                                    custom_bg = all_gene_ids, # make sure this is all gene ids 
                                    multi_query = TRUE)
```


can you plot the results for different clusters together ?

```{r}
gostplot(multi_gostquery_results_obj, capped = TRUE, interactive = TRUE)
```



