---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "09/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?

```{r}
ggplot(diamonds)
```

> Answer:
## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:
## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
  
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
 ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point()
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

- Assign that last plot to an object called `obds_diamonds`.

```{r}
obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
print(obds_diamonds)
```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:
# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight)) + geom_point()
```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour= Diet)) + geom_point()
```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour= Diet)) + geom_point() + geom_smooth(method="lm")
```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:
- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x=Time, y=weight, colour= Diet)) + geom_point() + geom_smooth(method="lm") + facet_wrap(~Diet, nrow=2)
```

- Assign that last plot to an object called `obds_chickweight`.

```{r}
obds_chickwight <- ggplot(ChickWeight, aes(x=Time, y=weight, colour= Diet)) + geom_point() + geom_smooth(method="lm") + facet_wrap(~Diet, nrow=2)
```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
?ggplot2::msleep
```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x=order)) + geom_bar()
```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order)) + geom_bar() +theme(axis.text.x = element_text(angle=90,size = 9, hjust = 1, vjust = 0.5))
```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order)) + geom_bar() +theme(axis.text.x = element_text(angle=90,size = 9, hjust = 1, vjust = 0.5)) + labs(x="Taxonomic Order", y="Number of Animals", title = "MSleep Data Set")
```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

```{r}
ggplot(msleep, aes(x=order, fill = genus)) + geom_bar() +theme(axis.text.x = element_text(angle=90,size = 9, hjust = 1, vjust = 0.5)) + labs(x="Taxonomic Order", y="Number of Animals", title = "MSleep Data Set")
```

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order, fill = genus)) + geom_bar() +theme(axis.text.x = element_text(angle=90,size = 9, hjust = 1, vjust = 0.5), legend.key.size =unit(2, "mm")) + labs(x="Taxonomic Order", y="Number of Animals", title = "MSleep Data Set") 
```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order, fill = genus)) + geom_bar() +theme(axis.text.x = element_text(angle=90,size = 9, hjust = 1, vjust = 0.5), legend.key.size =unit(2, "mm")) + labs(x="Taxonomic Order", y="Number of Animals", title = "MSleep Data Set") + guides(fill = guide_legend(ncol = 3))
```

- Add a contour of thin black lines to the bars.

```{r}
ggplot(msleep, aes(x=order, fill = genus)) + geom_bar(colour = "black", size = 0.1) +theme(axis.text.x = element_text(angle=90,size = 9, hjust = 1, vjust = 0.5), legend.key.size =unit(2, "mm")) + labs(x="Taxonomic Order", y="Number of Animals", title = "MSleep Data Set") + guides(fill = guide_legend(ncol = 3))
```

- Assign that last plot to an object called `obds_msleep`.

```{r}
obds_msleep <- ggplot(msleep, aes(x=order, fill = genus)) + geom_bar(colour = "black", size = 0.1) +theme(axis.text.x = element_text(angle=90,size = 9, hjust = 1, vjust = 0.5), legend.key.size =unit(2, "mm")) + labs(x="Taxonomic Order", y="Number of Animals", title = "MSleep Data Set") + guides(fill = guide_legend(ncol = 3))
```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r}
library(cowplot)
first_row <- cowplot::plot_grid(obds_diamonds, obds_chickwight, labels = c("OBDS_diamonds", "OBDS_chickweight"), ncol =2, nrow = 1)
second_row <- cowplot::plot_grid(obds_msleep, ncol = 1, nrow =1)
super_plot <- cowplot::plot_grid(first_row, second_row, ncol = 1, nrow =2)
super_plot
```

Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)
pathwork_plot <- (obds_diamonds + obds_chickwight) / (obds_msleep)
```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}
library(gridExtra)
lay <- rbind(c(1,2), c(3,3))
gridextra_plot <- gridExtra::grid.arrange(obds_diamonds, obds_chickwight, obds_msleep, layout_matrix=lay)
```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave("super_plot.pdf", plot = super_plot, width = 10, height = 10)
ggsave("patchwork_plot.pdf", plot = pathwork_plot, width = 10, height = 10)
pdf(file = "gridextra_plot.pdf", width = 10, height = 10)
gridExtra::grid.arrange(gridextra_plot)
dev.off()
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}
ggplot(ggplot2::mpg, aes(x=cty, y=hwy, color = year)) +
  geom_point(size=0.5)+
  geom_abline(intercept = 0, slope = 1, linetype ="dashed", size = 0.5)+
  facet_wrap(~ manufacturer) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), strip.text = element_text(size=12), plot.title = element_text(hjust = 0.5)) +
  labs(y="Highway petrol usage", x="City petrol usage", title="Main Title", subtitle = "Plot Subtitle")
```

##Extra
```{R}
df1 <- data.frame(col1 = c(1,2,3), col2=c(6,5,4))
df2 <- data.frame(column1 = c(1.5,1.75), column2 =c(3.5,4))
df3 <- data.frame(variable1 = runif(100,min = 1, max = 3))
ggplot() +
  geom_point(aes(x = col1, y = col2), df1, color = "blue") +
geom_point(aes(x = column1, y = column2), df2, color = "red")
```

##Extra no 2
```{R}
df1 <- data.frame(col1 = c(1, 2, 3), col2 = c(6, 5, 4))
df2 <- data.frame(column1 = c(1.5, 1.75), column2 = c(3.5, 4))
df3 <- data.frame(variable1 = runif(100, min = 1, max = 3))
ggplot() +
geom_histogram(aes(x = variable1), df3, alpha = 0.5, bins = 100) +
geom_point(aes(x = col1, y = col2), df1, color = "blue") +
geom_point(aes(x = column1, y = column2), df2, color = "red")
```
